Article analysis

### 2017-10-07 Article analysis notebook.Rmd
Friday - error in dataset

#Step 1. Load packages and data
Before I run any parts of the analysis, I load all the packages I use in this file
```{r load libraries}
require(dplyr)
require(ggplot2)
require(stringr)
require(plotrix) #to calculate standard error
```
First in excel, I deleted two rows of non-table data, i.e. title, and a note, and saved the file (S2_Table) as a CSV in the project directory. Then I read it into R and converted it to a table using dplyr. I trimed white space at beginnings and ends of things where there shouldn't be any, and kept strings as is to avoid issues later down the line. 
```{r readfile arch}
iso_total<-read.csv("S2_Table.csv", header=TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
iso_total<-tbl_df(iso_total)
head(iso_total)
str(iso_total)
```
```{r readfile modern}
modern<-read.csv("S1_Table.csv", header=TRUE, skip=3, strip.white=TRUE, stringsAsFactors = FALSE) #here I used skip=3 to skip the first 3 rows before the data started
modern<-tbl_df(modern)
head(modern)
str(modern)
```

To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number. 
```{r rename}
names(iso_total)[12]<-"C13ap" #I used names(iso_total) to get the position of these for renaming
names(iso_total)[13]<-"O18ap"
names(modern)[10]<-"modC13"
names(modern)[11]<-"modO18"
```

@@

Here, I used the filter function to select only those rows where C/P and IR-SF fell within the expected ranges, and saved this as a new dataset "iso". I also needed to make sure the rows with NA values were still included, since only a random sample was analysed. I did this using the | notation which functions like "OR". Since the FTIR analysis returned both C/P and IR-SF values, I only need to include one is.na function here. I also excluded an outlier with an extremely low delta O18 that they excluded by converting this number to NA, as the samples delta C13 value was still incorperated in the analysis.
```{r filter out diagenisis}
#to filter out only Lepus and Sylvilagus genus, I first needed to define the column as a factor
iso_total$Genus<-as.factor(iso_total$Genus)
iso<-filter(iso_total, (Genus=="Lepus" | Genus=="Sylvilagus") & between (C.P, 0.1, 0.5)&between (IR.SF, 2.0, 4.0) | (is.na(IR.SF))) #brackets around the Genus filter were important here, otherwise the | worked on everything that came after
iso$O18ap[iso$O18ap==11.9]<-NA #couldn't figure out how to get this to work with dplyr or chaining
iso            
```

"δ13Capatite = -7.8 ± 2.4‰(N = 114, 1 S.D.) and δ18Oapatite = 26.0‰(N = 113, 1 S.D.). The modern reference specimens exhibited
mean values of δ13Capatite = -12.4 ± 1.8‰(N = 13, 1 S.D.) and δ18Oapatite = 26.1‰± 1.4 (N = 13, 1 S.D.)."(Somerville et al. 2016:10). *I also note that they should have used "n" here*

@@

```{r sum_table}
my_sum <- 
  list("Archaeological samples" = 
         list("δC13 mean" = ~ mean(C13ap, na.rm=TRUE),
              "δC13 sd" = ~ sd(C13ap, na.rm=TRUE),
              "δO18 mean" = ~ mean(O18ap, na.rm=TRUE),
              "δO18 sd" = ~ sd(O18ap, na.rm=TRUE)),
       "Modern samples" = 
         list("δC13 mean" = ~ mean(modC13, na.rm=TRUE),
              "δC13 sd" = ~ sd(modC13, na.rm=TRUE),
              "δO18 mean" = ~ mean(modO18, na.rm=TRUE),
              "δO18 sd" = ~ sd(modO18, na.rm=TRUE))
  )

summary_table(iso, my_sum[1])
summary_table(modern, my_sum[2])
```

```{r summary}
sum(!is.na(iso$C13ap))#the ! means it will return the number that is not NA
sum(!is.na(iso$O18ap))
arch_sum<-summarize(iso, meanC13=mean(C13ap, na.rm=TRUE), meanO18=mean(O18ap, na.rm=TRUE), sdC13=sd(C13ap, na.rm=TRUE), sdO18=sd(O18ap, na.rm=TRUE))%>%print()

mod_sum<-summarize(modern, meanC13=mean(modC13, na.rm=TRUE), meanO18=mean(modO18, na.rm=TRUE), seC13=std.error(modC13, na.rm=TRUE), sdO18=sd(modO18, na.rm=TRUE))%>%print()

```

##Step 3

>>>>

#Step 1. Load packages and data
Before I run any parts of the analysis, I load all the packages I use in this file
```{r load libraries}
require(dplyr)
require(ggplot2)
require(stringr)
require(plotrix) #to calculate standard error
require(qwraps2)#for summary table
require(tidyr)
```
First in excel, I deleted two rows of non-table data, i.e. title, and a note, and saved the file (S2_Table) as a CSV in the project directory. Then I read it into R and converted it to a table using dplyr. I trimed white space at beginnings and ends of things where there shouldn't be any, and kept strings as is to avoid issues later down the line. 
```{r readfile arch}
iso_total<-read.csv("S2_Table.csv", header=TRUE, stringsAsFactors = FALSE, strip.white=TRUE, na.strings = c("", "NA")) # there were some rows of blanks being included so I used this last bit to convert them to NA
iso_total<-tbl_df(iso_total)
head(iso_total)
str(iso_total)
```
```{r readfile modern}
modern<-read.csv("S1_Table.csv", header=TRUE, skip=3, strip.white=TRUE, stringsAsFactors = FALSE, na.strings = c("", "NA")) #here I used skip=3 to skip the first 3 rows before the data started
modern<-tbl_df(modern)
head(modern)
str(modern)
```

To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number. 
```{r rename}
names(iso_total)[12]<-"C13ap" #I used names(iso_total) to get the position of these for renaming
names(iso_total)[13]<-"O18ap"
names(iso_total)[1]<-"lab_n"
names(modern)[10]<-"modC13"
names(modern)[11]<-"modO18"
```

@@

Here, I used the filter function to select only those rows where C/P and IR-SF fell within the expected ranges, and saved this as a new dataset "iso". I also needed to make sure the rows with NA values were still included, since only a random sample was analysed. I did this using the | notation which functions like "OR". Since the FTIR analysis returned both C/P and IR-SF values, I only need to include one is.na function here. I also excluded an outlier with an extremely low delta O18 that they excluded by converting this number to NA, as the samples delta C13 value was still incorperated in the analysis.
```{r filter out diagenisis}
#to filter out only Lepus and Sylvilagus genus, I first needed to define the column as a factor
iso_total$Genus<-as.factor(iso_total$Genus)
iso<-filter(iso_total, (Genus=="Lepus" | Genus=="Sylvilagus") & between (C.P, 0.1, 0.5)&between (IR.SF, 2.0, 4.0) | (is.na(IR.SF))) #brackets around the Genus filter were important here, otherwise the | worked on everything that came after
iso$O18ap[iso$O18ap==11.9]<-NA #couldn't figure out how to get this to work with dplyr or chaining
iso            
```
```{r}
iso<-filter(iso, !(is.na(iso$Genus))) #removes empty rows
```

**Descriptive statistics from the article**
Archaological specimens: δ13Capatite = -7.8 ± 2.4‰(N = 114, 1 S.D.) and δ18Oapatite = 26.0‰(N = 113, 1 S.D.). 
Modern reference specimens: δ13Capatite = -12.4 ± 1.8‰(N = 13, 1 S.D.) and δ18Oapatite = 26.1‰± 1.4 (N = 13, 1 S.D.)."(Somerville et al. 2016:10). *I also note that they should have used "n" here*

@@

```{r sum_table}
my_sum <- 
  list("Archaeological samples" = 
        round (c("δC13 n" = sum(!is.na(iso$C13ap)),
              "δC13 mean" =  mean(iso$C13ap, na.rm=TRUE),
              "δC13 sd" =  sd(iso$C13ap, na.rm=TRUE),
              "δO18 n" = sum(!is.na(iso$O18ap)),
              "δO18 mean" =  mean(iso$O18ap, na.rm=TRUE),
              "δO18 sd" =  sd(iso$O18ap, na.rm=TRUE)),2),
       "Modern samples" = 
         round(c("δC13 n" = sum(!is.na(modern$modC13)),
              "δC13 mean" =  mean(modern$modC13, na.rm=TRUE),
              "δC13 sd" =  sd(modern$modC13, na.rm=TRUE),
              "δO18 n" = sum(!is.na(modern$modO18)),
              "δO18 mean" =  mean(modern$modO18, na.rm=TRUE),
              "δO18 sd" =  sd(modern$modO18, na.rm=TRUE)),2)
       )

my_sum
```

#Step 3 "Independent samples t-tests were conducted to compare mean stable isotope ratios of the two genera (Lepus and Sylvilagus) across the archaeological bone sample as a whole. No significant difference between the genera were found in δ13Capatite values (t[110] = -0.806, p = 0.422) or δ18Oapatite values (t[110] = 0.428, p = 0.669)."

First, I will check that the data fits a normal distribution by comparing it to the qnorm and qqline plots as well as histogram.

**Carbon Apatite Results**
*Histogram*
```{r}
a_hist<-ggplot(iso, aes(C13ap))
a_hist+geom_histogram(bins= 15)+ facet_wrap(~Genus, scales="free")
```
Sylvilagus looks approximately normal while Lepus it is hard to tell. Next I compare using qqplots.
*QQplots*
```{r qqplots}
LC<-iso$C13ap[iso$Genus=="Lepus"]
SC<-iso$C13ap[iso$Genus=="Sylvilagus"]
qqnorm(LC,xlab="Theoretical",main="Lepus")
  qqline(LC)

  qqnorm(SC,xlab="Theoretical",main="Sylvilagus")
  qqline(SC)
```
Again, Lepus fits this "OK" while Sylvilagus looks pretty good. 
*Test of equal variance*
```{r}
var.test(LC, SC)
```

This suggests that the variance of the two samples is comparable and thus a two sample T test can be used.First I report the tests for the oxygen data and then provide both t-tests together with a discussion.

**Oxygen Apatite Results**
*Histogram*
```{r}
a_hist<-ggplot(iso, aes(O18ap))
a_hist+geom_histogram(bins=15)+ facet_wrap(~Genus, scales="free")
```
Both look like they could be approximately normal
*QQplots*
```{r qqplots}
LO<-iso$O18ap[iso$Genus=="Lepus"]
SO<-iso$O18ap[iso$Genus=="Sylvilagus"]
qqnorm(LO,xlab="Theoretical",main="Lepus")
  qqline(LO)

  qqnorm(SO,xlab="Theoretical",main="Sylvilagus")
  qqline(SO)
```
Here, Lepus fits the normal distrbution well, while Sylvilagus is divergent. Looking at the histgram again, it could poentitally be bimodal. 
*Test of equal variance*
```{r}
var.test(LO, SO)
```
*Carbon apatite two sample T-test*
```{r}
t.test(LC, SC, var.equal =TRUE, alternative = "two.sided")
```

*Oxygen apatite two sample T-test*
```{r}
t.test(SO, LO, var.equal =TRUE, alternative = "two.sided")
```
*Summary:*
Carbon apatite t-sample t-test comparing Lepus and Sylvilagus: t = 0.5185, df = 112, p-value = 0.6051
Oxygen apatite t-sample t-test comparing Lepus and Sylvilagus: t = -0.83133, df = 111, p-value = 0.4076
*Values reported in Somerville et al. (2016:10)*
δ13Capatite values (t[110] = -0.806, p = 0.422) 
δ18Oapatite values (t[110] = 0.428, p = 0.669).

Both of these tests returned different test statistics, degrees of freedom and p-values to those reported by Somerville et al. (2016). I tried multiple variations of the t-test in R, from Welsh t-test and one sided t-tests to one sided Welsh t-tests and none of the tests produced a similar result to those reported by Somerville. In fact, the two sample t-test for equal varience that I ran for both carbon and oxygen data matches most closely with the result reported for the opposite isotopic data. i.e. my oxygen looks most similar to their carbon. I have a number of hypothesis for why these tests may vary: the SPSS test is vastly different to the R t-tests, they are excluding samples from the test and not reporting it, as indicated by the difference in df between our tests, or there was an error in their reporting of the data. That said, the outcome of all tests is the same, there is no significant difference in the means of the samples. I am also not sure about the utility of this test in general, given that the article is looking for differences within a single species as well to test for outliers in the diet of these animals.

#Step 4 
"We also evaluated stable isotope ratios between the genera and between individual site locations using a two-way ANOVA (factorial ANOVA ith two factors), testing for the main effects of genus and context, and for the genuscontext interaction."

```{r}
carbon<-lm(iso$C13ap~iso$Genus*iso$Site)
anova(carbon)
```
```{r}
oxygen<-lm(iso$O18ap~iso$Genus*iso$Site)
anova(oxygen)
```

### 2017-10-07 Article analysis.Rmd
Friday - error in dataset

>>>>

```{r summary}
sum(!is.na(iso$C13ap))#the ! means it will return the number that is not NA
sum(!is.na(iso$O18ap))
arch_sum<-summarize(iso, meanC13=mean(C13ap, na.rm=TRUE), meanO18=mean(O18ap, na.rm=TRUE), sdC13=sd(C13ap, na.rm=TRUE), sdO18=sd(O18ap, na.rm=TRUE))%>%print()

mod_sum<-summarize(modern, meanC13=mean(modC13, na.rm=TRUE), meanO18=mean(modO18, na.rm=TRUE), seC13=std.error(modC13, na.rm=TRUE), sdO18=sd(modO18, na.rm=TRUE))%>%print()

These results also differ from what was presented. So I will test to see if they removed outliers prior to t-test. I will use a box plot to identify outliers
```{r}
boxplot(SC, ylim = c(), main = "dC13 ap", xlab = "Sylvilagus")
boxplot(LC, ylim = c(), main = "dC13 ap", xlab = "Lepus")
boxplot(SO, ylim = c(), main = "dO18 ap", xlab = "Sylvilagus")
boxplot(LO, ylim = c(), main = "dO18 ap", xlab = "Lepus")
```
